## **R Libraries: ggplot2**

The R library `ggplot2` is a super useful tool we can use to create visualizations and explore our data.

```{bash}
install.packages(tidyverse)
library(tidyverse)
# had to manually type each into the console or search for and instaclick on in 'packages'
```

More things we can do to get more information on R and what we're doing:

```{bash}
help()
```

We'll need a few more packages for our exploration usage, too. We can install them all at once by using the `c()` function, which creates a list. We'll open these packages with the `library()` function later.

```{bash}
install.packages(c("nycflights13", "gapminder", "Lahman"))
# had to manually type each into the console
# i.e. install.packages("Lahman")
```

Built in to our tidyverse are several data sets that we can play around with. In the below code block, we create a variable `myplot` and save the `mpg` database to it.

```{r}
# mpg -> subset of car fuel economy data from the EPA
myplot <- mpg
myplot
# datatype ; a 'tibble' ; like a more frustrating python dataframe, they complain about things that dataframes don't, however it is nice because it complains up front so we can fix things instead of doing it during a presentation in front of our bosses

# excellent tool for viewing whole dataset in a spreadsheet; supports sorting by column and searching
View(myplot)
```

**Something you should be aware of and have open (perhaps on a second monitor, or you can print them out) is the ggplot2 cheat sheet!** <https://github.com/rstudio/cheatsheets/blob/main/data-visualization-2.1.pdf>

```{r}
# 'hey I'm gonna use ggplot, I'm going to make a plot and pass in the mpg dataset for my data
# geom_point() creates scatterplots
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

# engine displacement -> how big the engine is
# hwy -> mileage

# smaller the engine, the better the gas mileage
```

```{r}
# motor trend dataset ; design and performance on 32 cars
mtcars 

View(mtcars)
```

```{r}
# can do on one line or split on the +
# add in color scheme by a column
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = class))

# whoaaaaa it handles everything for you and spits out an incredible pretty graph
```

```{r}
# color by conditional statement
ggplot(data = mpg) + geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

```{r}
# facet_wrap() separates subplots by a column grouping
# could have done color = rgb(0.7,1,0.3)
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) +
  facet_wrap(~ class, nrow = 2)

# can also add aes(x=,y=,color=,shape = class) for interesting effect (only 6 default symbols however, can specify a manual shape)
```

```{r}
# 
# looking at miles per gallon x engine size for different drive train types
# facet_grid(drv ~ .) facet_grid is a good type comparison ; 'drv ~ .' = use all drive trains
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = drv)) +
  facet_grid(drv ~ .)
```

```{r}
# geom_smooth is a great way of showing area/direction of different subsets in a graph
# aes(linetype)
# two plots on top of each other
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) +
  geom_point(mapping = aes(x = displ, y = hwy, color = drv))

# without geom_smooth area shading, gets a bit confusing in the middle
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, linetype = drv)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(se = FALSE)
```

```{r}
# discrete graphs look for an attribute of our data
# diamonds dataset came with ggplot
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

Whoa, where'd count come from? **Count is not a variable or attribute of the variable diamonds!** Count comes automatically with a lot of graphs as an attribute. You can verify what I'm saying with `?diamonds` in the console. It shows up because this is a discrete data set, rather than the continuous data set that we have been using.

Any attribute of a data set that is algorithmically calculated is called a *stat*, which is short for a statistical transformation. Many of the **geom** functions have stats built in, and many stats display geoms. For example, the above code block used a geom, but this one uses a stat, and results in an identical chart:

```{r}
# shows that stat_count is built into geom_bar
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

Let's add color here!

```{r}
# aes(color) = for bar graphs only outlines, not easy to see
# aes(fill) = much betterrrrr
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut))
```

What if we add to a different variable than cut?

```{r}
# position = "fill" good option for classic pie chart
# theme()
# labs() for labels, here we are getting rid of x/y labels
# coord_polar() used for pie charts
pie = ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity),
           position = "dodge"
  ) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

pie + coord_polar()

# can see which clarities have the best cuts

# sidenote: if you have a lot of datapoints but they're very similar, use geom_jitter() and it will move points around to show the minute differences and densities
```

**For your consideration...** Is the above a good graph?

```{r}
# little too small to see clearly
```

## Homework

```{r}
# Find a builtin data set for students to explore; use their cheat sheet from stats to label continuous and discrete data; Use graphs to tell me things about 3 datsets

# help -> ggplot2-package -> look for datasets that came with it -> diamonds, mpg, txhousing

# nycflights13 has sooooo much data and data subsets accessed by nycflights::subset (will fill in options, there are 5)

View(nycflights13::airlines)
View(nycflights13::airports)
View(nycflights13::flights)
  # discrete: carriers, tailnumbers
View(nycflights13::planes)
View(nycflights13::weather)

```

```{r}
# airplane speed based on number of engines and type
ggplot(data = nycflights13::planes) +
  geom_col(mapping = aes(x = engines, y = speed, fill = engine))

# The fastest planes have 2 reciprocating engines
# Turbo-prop and turbo-shaft tend to be slow
# Planes with 1 or 4 reciprocating engines are slow
```

```{r}
# distribution of flight distances by carrier
# x = carrier, y = distance
# two discrete variables -> geom_count

ggplot(data = nycflights13::flights) +
  geom_count(mapping = aes(x = carrier, y = distance))

# Hawaiian goes long distance only; island
# US Airways is very concentrated in shirt and medium distance
# Virgin America is all medium distance; trans-Atlantic flights?
# United Air has the most spread distribution
```

```{r}
# wind speed by airport origin?
# one continuous variable, one discrete -> geom_col

ggplot(data = nycflights13::weather) +
  geom_col(mapping = aes(x = origin, y = wind_speed))

# JFK has the highest average wind speed out of the three! EWR the lowest
```
